---
title: "Simulate and fit closed and open spatial capture-recapture models"
date: "20/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      dpi = 300, 
                      message = FALSE, 
                      warning = FALSE, 
                      cache = TRUE)
library(tidyverse)
theme_set(theme_light())
```

## Motivation

I need to have in one place some code to simulate data and fit spatial capture-recapture models, for both closed and open populations. 

The reference book for spatial capture-capture (SCR) models is [Spatial Capture-Recapture](https://www.elsevier.com/books/spatial-capture-recapture/royle/978-0-12-405939-9) by Royle, Chandler, Sollmann and Gardner. Andy Royle has nice slides for [an introduction to SCR models](https://docs.google.com/viewer?a=v&pid=sites&srcid=ZGVmYXVsdGRvbWFpbnxzcGF0aWFsY2FwdHVyZXJlY2FwdHVyZXxneDo3MzMwZjQyYjA2M2MwMmNj), and a [video](https://www.youtube.com/watch?v=4HKFimATq9E). 

We will go the Bayesian way. Richard Chandler provides nice slides for Bayesian inference in [closed population SCR models](https://docs.google.com/viewer?a=v&pid=sites&srcid=ZGVmYXVsdGRvbWFpbnxzcGF0aWFsY2FwdHVyZXJlY2FwdHVyZXxneDozYzkyMjQwYTA3ZjYzNDFm), and [open population SCR models](https://docs.google.com/viewer?a=v&pid=sites&srcid=ZGVmYXVsdGRvbWFpbnxzcGF0aWFsY2FwdHVyZXJlY2FwdHVyZXxneDpkYWIyZGU3OTRjYTkwYzA).

To fit SCR models, we will use [`Nimble`](https://r-nimble.org/) developed by [Perry de Valpine and colleagues](https://r-nimble.org/more/about-us). We might implement [a few tricks by Daniel Turek](https://www.biorxiv.org/content/10.1101/2020.05.07.081182v1) to improve and reach convergence faster.

Load `Nimble`.
```{r}
library(nimble)
```

## Closed population spatial capture-recapture models

### Description

We use some code shared by Jose Jimenez on the [SCR Google group](https://groups.google.com/g/spatialcapturerecapture) and adapted by Perry de Valpine. A formal description of the model we consider is given in the references given above.

### Simulations

The traps.
```{r}
tr <- seq(15, 85, length = 10)
traps <- data.frame(X = rep(tr, each = length(tr)),
                    Y = rep(tr, times = length(tr))) # 100 coord. traps
```

Visualize.
```{r}
viz_traps <- traps %>% 
  ggplot(aes(x = X, y = Y)) +
  geom_point(pch = 3) + 
  xlim(0, 100) +
  ylim(0, 100)
viz_traps
```

Generate population.
```{r}
set.seed(10)
xlim <- c(0, 100)
ylim <- c(0, 100) # area 100 * 100 = 1e4
A <- (xlim[2] - xlim[1]) * (ylim[2] - ylim[1])/10000
A
mu <- 50 # density
N <- rpois(1, mu*A) # generate population
N 
```

Generate activity centers.
```{r}
s <- data.frame(s.x = runif(N, xlim[1], xlim[2]), 
                s.y = runif(N, ylim[1], ylim[2]))
```

Visualize.
```{r}
viz_traps_ac <- viz_traps + 
  geom_point(data = s, aes(x = s.x, y = s.y), pch = 16, color = "red")
viz_traps_ac
```

Generate detections.
```{r}
sigma <- 5
lambda0 <- 0.4
J <- nrow(traps) # nb of traps
K <- 5 # nb capture occasions
yy <- array(NA, c(N, J, K))
for(j in 1:J) {
  dist <- sqrt((traps$X[j] - s$s.x)^2 + (traps$Y[j] - s$s.y)^2)
  lambda <- lambda0 * exp(-dist^2 / (2 * sigma^2))
  for(k in 1:K) {
    yy[,j,k] <- rpois(N, lambda)
  }
}
n <- apply(yy, c(2,3), sum)
```


Plot detections.
```{r}
tot <- apply(n, 1, sum)
dat <- data.frame(traps, tot = tot)

viz_traps_ac +
  geom_point(data = dat, aes(x = X, y = Y, size = tot), alpha = 0.3) +
  scale_size(range = c(0, 20)) +
  labs(x = "",
       y = "",
       size = "# detections")
```


### Model fitting

Define the model.
```{r}
code <- nimbleCode({
  sigma ~ dunif(0, 10)
  lam0 ~ dunif(0, 5)
  psi ~ dbeta(1, 1)
  for(i in 1:M) {
    z[i] ~ dbern(psi)
    s[i,1] ~ dunif(xlim[1], xlim[2])
    s[i,2] ~ dunif(ylim[1], ylim[2])
    dist[i,1:J] <- (s[i,1] - X[1:J,1])^2 + (s[i,2] - X[1:J,2])^2
    lam[i,1:J] <- exp(-dist[i,1:J] / (2 * sigma^2)) * z[i]
  }
  for(j in 1:J){
    bigLambda[j] <- lam0 * sum(lam[1:M,j])
    for(k in 1:K) {
      n[j,k] ~ dpois(bigLambda[j])
    }
    }
  N <- sum(z[1:M])
})
```

Define constants, data and inits.
```{r}
M <- 200
constants <- list(M = M, 
                  K = K, 
                  J = J)
n1 <- apply(n, 1, sum)
data <- list(n = n, 
             X = traps, 
             xlim = xlim, 
             ylim = ylim)
s <- cbind(runif(M, xlim[1], xlim[2]), 
           runif(M, ylim[1], ylim[2]))
z <- rep(1, M)
inits <- list(sigma = 0.5, 
              lam0 = 0.1, 
              s = s, 
              z = z,
              psi = 0.5)
```


Build `R` model (not compiled yet).
```{r}
Rmodel <- nimbleModel(code = code, 
                      constants = constants, 
                      data = data, 
                      inits = inits)
```

Check whether the model is fully initialized. If you failed at providing initial values for some parameters (e.g. $\psi$), you'll get NAs.
```{r}
Rmodel$calculate()
```

Now compile the model in `C++`.
```{r}
Cmodel <- compileNimble(Rmodel)
```

The `R` and `C` models are exactly the same versions of the model.
```{r}
calculate(Cmodel)
```

You can simulate from prior.
```{r}
Cmodel$simulate('lam0')
calculate(Cmodel)
```

Specify MCMC.
```{r}
conf <- configureMCMC(Rmodel,
                      monitors = c("N", "lam0", "psi", "sigma"))
```

Build an executable MCMC.
```{r}
Rmcmc <- buildMCMC(conf)
```

Compile in `C++`. 
```{r}
Cmcmc <- compileNimble(Rmcmc, project = Cmodel)
```

Run compiled model (do not run the uncompiled model with `runMCMC(Rmcmc,100)`).
```{r}
samples <- runMCMC(Cmcmc,100)
```

Explore.
```{r}
dim(samples)
colnames(samples)
samplesSummary(samples)
```

Run compiled model. Takes 10-15 minutes on my machine.
```{r}
samplesList <- runMCMC(Cmcmc,
                   niter = 10000,
                   nburnin = 5000,
                   nchains = 2)
samples <- rbind(samplesList[[1]],
                 samplesList[[2]])
```

```{r}
str(samples)
```

Calculate ESS effective sample size. 
```{r}
library(coda)
apply(samples, 2, effectiveSize)
```

Produce trace an density plots. 
```{r}
library(basicMCMCplots)
chainsPlot(samplesList,
           var = c("N", "sigma", "lam0"))
```

Display summary stats. Compare to the values used to simulate data, in particuler $N = 50$, $\sigma = 5$ and $\lambda_0 = 0.4$.
```{r}
summary(samples)
```

## Open population spatial capture-recapture models

### Description

We use some code shared by Beth Gardner and colleagues in their paper [State space and movement specification in open population spatial capture–recapture models](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.4509). A formal description of the model we consider is given in the paper. We consider a model with constant activity centers.

### Simulations

Function to calculate the distance between multiple points.
```{r}
e2dist <- function(x, y){ 
  i <- sort(rep(1:nrow(y), nrow(x)))
  dvec <- sqrt((x[, 1] - y[i, 1])^2 + (x[, 2] - y[i, 2])^2)
  matrix(dvec, nrow = nrow(x), ncol = nrow(y), byrow = F)
}
```


```{r}
simJS.fn <- function(N, phi0, lam0, M, T, grid, xl, xu, yl, yu, sigma, K){ # M is total ever alive
  ntraps <- dim(grid)[1]
  nreps <- K
  lam0 <- rep(lam0, T)
  phi<- rep(phi0, T)
  pmat <- lam <- list()
  gamma <- NULL
  gamma[1] <- N/M
  sx <- runif(M, xl, xu)
  sy <- runif(M, yl, yu)
  z <- r <- al <- matrix(0, nrow = M, ncol = T)
  r[,1] <- rbinom(M, 1, gamma[1])
  z[,1] <- r[,1]
  for (t in 2:T){
    # survival
    surv <- rbinom(M, 1, z[,t-1] * phi[t])
    # recruitment
    al[,t] <- apply(matrix(z[,1:(t-1)], nrow = M, byrow = FALSE), 1, sum) > 0
    idx <- 1 - as.numeric(al[,t])
    gamma[t] <- (N - sum(surv)) / sum(idx)
    if (gamma[t] < 0) gamma[t] <- 0
    r[,t] <- rbinom(M, idx, gamma[t])
    z[,t] <- surv + r[,t]
  }
  S <- cbind(sx, sy)
  dmat <- e2dist(S, grid)
  psi<- exp(-(1 / (2 * sigma * sigma)) * dmat * dmat)
  for (t in 1:T){
    lam[[t]] <- lam0[t] * psi
    pmat[[t]] <- 1 - exp(-lam[[t]])
  }
  y <- array(0, dim = c(M, ntraps, T))
  for (t in 1:T){
    yfull <- array(0, dim = c(M, ntraps, K))
    for (i in 1:M){
      for (k in 1:K){
        yfull[i, 1:ntraps, k] <- rbinom(ntraps, 1, pmat[[t]][i,] * z[i,t])
      }
    }
    y[, 1:ntraps, t] <- apply(yfull, 1:2, sum)
  }
  ycapt <- y[which(rowSums(y[,,])>0), , ]
  list(y = ycapt, 
       z = z,
       r = r,
       gamma = gamma,
       N = apply(z, 2, sum),
       R = apply(r, 2, sum), 
       SX = sx, 
       SY = sy)
}
```

Set up the basic trap array in a 7x7 grid.
```{r}
gridx <- seq(-3, 3, 1)
grid <- as.matrix(expand.grid(gridx, gridx))
J <- dim(grid)[1]
```

Set the upper and lower x and y coordinates for the state space.
```{r}
xl <- -5
yl <- -5
xu <- 5
yu <- 5
```

Parameters specification.
```{r}
T <- K <- 5 # number of years / seasons
sigma <- 0.5
lam0 <- 0.5
N <- 40
M <- 150
phi0 <- 0.75
tau <- 0.5
Mc <- 150 # upper M for data augmentation
```

Simulate data.
```{r}
simdat <- simJS.fn(N = N,
                   phi0 = phi0,
                   lam0 = lam0,
                   M = Mc, 
                   T = T, 
                   grid = grid, 
                   xl = xl, 
                   xu = xu, 
                   yl = yl, 
                   yu = yu, 
                   sigma = sigma, 
                   K = K)
str(simdat)
```

Visualise traps.
```{r}
viz_traps <- grid %>% 
  as_tibble() %>%
  ggplot(aes(x = Var1, y = Var2)) +
  geom_point(pch = 3) +
  xlim(-5, 5) +
  ylim(-5, 5)
viz_traps
```

Add activity centers.
```{r}
viz_traps_ac <- viz_traps + 
  geom_point(data = data.frame(s.x = simdat$SX, s.y = simdat$SY), 
             aes(x = s.x, y = s.y), pch = 16, color = "red")
viz_traps_ac
```


### Model fitting

Define the model.
```{r}
code <- nimbleCode({
  # set priors for sigma2, lam0 (encounter rate), gamma, and phi
  sigma ~ dunif(0, 10)
  sigma2 <- sigma * sigma
  lam0 ~ dunif(0, 5)
  phi ~ dunif(0, 1) # survival
  for(t in 1:T){ # T = 10 years
    gamma[t] ~ dunif(0, 1) # recruitment
  }
  for (i in 1:M){ # loop over M individuals (includes the augmented data)
    ncaps[i] <- sum(z[i,1:T])
    alive[i] <- 1 - equals(ncaps[i],0)
    z[i,1] ~ dbin(gamma[1], 1)
    SX[i] ~ dunif(xl, xu) # set priors for the X and Y coordinates of each individual
    SY[i] ~ dunif(yl, yu)
    for(j in 1:J) { # loop over all traps of that year
      D2[i,j] <- pow(SX[i] - trapmat[j,1], 2) + pow(SY[i] - trapmat[j,2], 2)
      g[i,j] <- lam0 * exp(-D2[i,j] / (2 * sigma2))
      pmean[i,j] <- 1 - exp(-g[i,j])
      for (t in 1:T){
      tmp[t,i,j] <- z[i,t] * pmean[i,j]
      y[t,i,j] ~ dbin(tmp[t,i,j], K) # K is the number of days a trap was operational
      }
    }
  a[i,1] <- (1 - z[i,1])
  a[i,2] <- (1-z[i,1]) * (1-z[i,2]) 
  a[i,3] <- (1-z[i,1]) * (1-z[i,2]) * (1-z[i,3]) 
  a[i,4] <- (1-z[i,1]) * (1-z[i,2]) * (1-z[i,3]) * (1-z[i,4]) 

  gammatmp[i,2] <- gamma[2] * a[i,1]
  mu[i,2] <- (phi * z[i,1]) + gammatmp[i,2]
  z[i,2] ~ dbern(mu[i,2])
  
  gammatmp[i,3] <- gamma[3] * a[i,2]
  mu[i,3] <- (phi * z[i,2]) + gammatmp[i,3]
  z[i,3] ~ dbern(mu[i,3])

  gammatmp[i,4] <- gamma[4] * a[i,3]
  mu[i,4] <- (phi * z[i,3]) + gammatmp[i,4]
  z[i,4] ~ dbern(mu[i,4])

  gammatmp[i,5] <- gamma[5] * a[i,4]
  mu[i,5] <- (phi * z[i,4]) + gammatmp[i,5]
  z[i,5] ~ dbern(mu[i,5])

  R[i,1] <- z[i,1]
  R[i,2] <- (1 - z[i,1]) * z[i,2]
  R[i,3] <- (1 - z[i,1]) * (1 - z[i,2]) * z[i,3]
  R[i,4] <- (1 - z[i,1]) * (1 - z[i,2]) * (1 - z[i,3]) * z[i,4]
  R[i,5] <- (1 - z[i,1]) * (1 - z[i,2]) * (1 - z[i,3]) * (1 - z[i,4]) * z[i,5]
  }
  
  N1 <- sum(z[1:M,1])
  N2 <- sum(z[1:M,2])
  N3 <- sum(z[1:M,3])
  N4 <- sum(z[1:M,4])
  N5 <- sum(z[1:M,5])
  Nalive <- sum(alive[1:M])
  
  R1 <- sum(R[1:M,1])
  R2 <- sum(R[1:M,2])
  R3 <- sum(R[1:M,3])
  R4 <- sum(R[1:M,4])
  R5 <- sum(R[1:M,5])
})
```


Function to create initial coordinates of activity centers for each individual.
```{r}
Sin <- function(T = T, M = M, xl = xl, xu = xu, yl = yl, yu = yu, ntot = ntot){
  SX <- SY <- matrix(NA, nrow = M, ncol = 1)
  for(i in 1:M){
    for (t in 1:T){
      SX[i] <- runif(1, xl, xu)
      SY[i] <- runif(1, yl, yu)
      traps <- which(dataug[i,,t] > 0)
      if(length(traps) > 0){
        SX[i] <- mean(grid[traps,1])
        SY[i] <- mean(grid[traps,2])
      }
    }
  }
return(list(SX, SY))
}
```


Data, inits and parameters to monitored. 
```{r}
ntot <- dim(simdat$y)[1] # total ever observed in this simulated dataset
# add Mc-ntot zero encounter histories (data augmentation)
dataug <- array(0, dim = c(Mc, J, T))
dataug[1:ntot, , ] <- simdat$y
dataugTMJ <- aperm(dataug, c(3,1,2))
# create intial values for z state
zinit <- matrix(0,nrow = Mc, ncol = T)
zinit[1:ntot,] <- 1

constants <- list(M = Mc, 
                  K = K, 
                  J = J,
                  T = T)

data <- list(y = dataugTMJ, 
             xl = xl, 
             xu = xu, 
             yl = yl, 
             yu = yu, 
             trapmat = as.matrix(grid))

inits <- list(phi = runif(1), 
       gamma = runif(T, 0, 1),
       sigma = runif(1, 1, 2),
       z = zinit,
       lam0 = runif(1), 
       SY = as.vector(Sin(T = T, M = Mc, xl = xl, xu = xu, yl = yl, yu = yu, ntot = ntot)[[2]]),
       SX = as.vector(Sin(T = T, M = Mc, xl = xl, xu = xu, yl = yl, yu = yu, ntot = ntot)[[1]]))
```


Build `R` model (not compiled yet).
```{r}
Rmodel <- nimbleModel(code = code, 
                      constants = constants, 
                      data = data, 
                      inits = inits)
```

Check whether the model is fully initialized.
```{r}
Rmodel$calculate()
```

Now compile the model in `C++`.
```{r}
Cmodel <- compileNimble(Rmodel)
```

The `R` and `C` models are exactly the same versions of the model.
```{r}
calculate(Cmodel)
```

Specify MCMC.
```{r}
conf <- configureMCMC(Rmodel,
                      monitors = c("N1", "N2", "N3", "N4", "N5",
                                   "R1", "R2", "R3", "R4", "R5",
                                   "lam0", "phi", "sigma", "gamma"))
```

Build an executable MCMC.
```{r}
Rmcmc <- buildMCMC(conf)
```

Compile in `C++`. 
```{r}
Cmcmc <- compileNimble(Rmcmc, project = Cmodel)
```

Run compiled model (do not run the uncompiled model with `runMCMC(Rmcmc,100)`).
```{r}
samples <- runMCMC(Cmcmc,100)
```

Explore.
```{r}
dim(samples)
colnames(samples)
samplesSummary(samples)
```

Run compiled model. Takes 10-15 minutes on my machine.
```{r}
samplesList <- runMCMC(Cmcmc,
                   niter = 10000,
                   nburnin = 5000,
                   nchains = 2)
samples <- rbind(samplesList[[1]],
                 samplesList[[2]])
```

```{r}
str(samples)
```

Calculate ESS effective sample size.
```{r}
library(coda)
apply(samples, 2, effectiveSize)
```

Produce trace an density plots. 
```{r}
library(basicMCMCplots)
chainsPlot(samplesList,
           var = c("N1", "N2", "N3", "N4", "N5"))
```

```{r}
chainsPlot(samplesList,
           var = c("R1", "R2", "R3", "R4", "R5"))
```

```{r}
chainsPlot(samplesList,
           var = c("sigma", "phi", "lam0"))
```


Display summary stats. Compare to the values used to simulate data, in particuler $N = 40$, $\sigma = 0.5$, $\phi = 0.75$ and $\lambda_0 = 0.5$.
```{r}
samplesSummary(samples)
```

Plot estimated pop size. 
```{r}
res <- samplesSummary(samples)
res <- cbind(param = rownames(res), res)
res %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(mean = round(as.numeric(mean)),
         low = round(as.numeric(x95_percent_ci_low)),
         up = round(as.numeric(x95_percent_ci_upp))) %>%
  filter(str_detect(param, "N")) %>%
  mutate(year = row_number()) %>%
  ggplot() + 
  aes(x = year, y = mean) + 
  geom_point() + 
  geom_errorbar(aes(ymin = low, ymax = up), width=.1)
```


## To-do list

* Add formal description of the models.

* Add code to plot map of estimated density and activity centers.

* Improve convergence.

* Provide Jags implementation (?).

* Provide secr and oSCR implementation for max likelihood counterpart (?).

